---
title: "Astrocyte Morphology Analysis"
author: "Habib Lab"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

This notebook analyzes **AstroResultsAstrocytes.csv** across all five time points (`ctrl`, `4h`, `24h`, `72h`, `7d`). Cell morphology/intensity uses **10–90% trimmed means** per image. Skeleton metrics are **means per astrocyte**.

```{r libraries}
library(tidyverse)
library(pheatmap)
source("astro_analysis_helpers.R")
```

```{r configuration}
DATA_PATH <- "C:/Users/talku/OneDrive/Habib Lab/CellProfiler_Results/AstroResultsAstrocytes.csv"
OUTPUT_ROOT <- "C:/Users/talku/OneDrive/Documents/HabibLab/output/all_times"

TIME_ORDER <- c("ctrl", "4h", "24h", "72h", "7d")
USE_TRIMMED_MEANS <- TRUE
TRIM_LOWER <- 0.10
TRIM_UPPER <- 0.90

OUT <- make_output_folders(OUTPUT_ROOT, c("heatmaps", "stats", "boxplots"))
TIME_COLORS <- get_time_colors(TIME_ORDER)
```

```{r load-and-process}
cells_data <- load_astro_cells(DATA_PATH)
features <- identify_feature_columns(cells_data)
image_metadata <- build_image_metadata(cells_data, TIME_ORDER)
TIME_COLORS <- get_time_colors(levels(image_metadata$Time))

image_means <- build_image_means(
  cells_data = cells_data,
  cell_features = features$cell,
  skeleton_features = features$skeleton,
  image_metadata = image_metadata,
  use_trimmed = USE_TRIMMED_MEANS,
  trim_lower = TRIM_LOWER,
  trim_upper = TRIM_UPPER
)

mean_matrix <- feature_matrix(image_means, features$all)
mean_matrix_core <- mean_matrix[, features$core, drop = FALSE]
row_annotation <- data.frame(
  Time = image_means$Time,
  row.names = rownames(mean_matrix)
)
ann_colors <- list(Time = TIME_COLORS[levels(row_annotation$Time)])
mean_method_label <- paste0(
  TRIM_LOWER * 100, "–", TRIM_UPPER * 100,
  "% trim + skeleton per cell, z-scored per feature"
)

cat("Images:", nrow(image_metadata), "\n")
cat("All features:", length(features$all), "\n")
cat("Core features:", length(features$core), "\n")
```

# Heatmaps

```{r heatmap-all-features}
plot_image_heatmap(
  scaled_mat = scale_features(mean_matrix),
  row_ann = row_annotation,
  ann_colors = ann_colors,
  title = paste0("All features\n(", mean_method_label, ")"),
  file = file.path(OUT$heatmaps, "all_features.png")
)
```

```{r heatmap-core-features}
plot_image_heatmap(
  scaled_mat = scale_features(mean_matrix_core),
  row_ann = row_annotation,
  ann_colors = ann_colors,
  title = paste0("Core features\n(", mean_method_label, ")"),
  file = file.path(OUT$heatmaps, "core_features.png")
)
```

```{r heatmap-cosine-all}
plot_cosine_heatmap(
  cosine_mat = cosine_similarity_matrix(mean_matrix),
  title = "Cosine similarity — all features",
  file = file.path(OUT$heatmaps, "cosine_all.png")
)
```

```{r heatmap-cosine-core}
plot_cosine_heatmap(
  cosine_mat = cosine_similarity_matrix(mean_matrix_core),
  title = "Cosine similarity — core features",
  file = file.path(OUT$heatmaps, "cosine_core.png")
)
```

# Feature ranking

Kruskal–Wallis on image-level values; features ranked by ε² = `(H − k + 1) / (N − k)`.

```{r time-difference-features}
time_feature_stats <- compute_time_feature_stats(
  image_means = image_means,
  feature_cols = features$all,
  use_trimmed = USE_TRIMMED_MEANS,
  trim_lower = TRIM_LOWER,
  trim_upper = TRIM_UPPER
)

core_time_stats <- compute_time_feature_stats(
  image_means = image_means,
  feature_cols = features$core,
  use_trimmed = USE_TRIMMED_MEANS,
  trim_lower = TRIM_LOWER,
  trim_upper = TRIM_UPPER
)
```

```{r plot-top25-kruskal}
plot_time_feature_ranking(
  time_feature_stats = time_feature_stats,
  time_colors = TIME_COLORS,
  n_features = 25,
  file = file.path(OUT$stats, "top25_kruskal.png"),
  title = "Top 25 features — Kruskal–Wallis ε²",
  subtitle = paste0(
    "Image-level values (", TRIM_LOWER * 100, "–", TRIM_UPPER * 100,
    "% trim + skeleton per cell); dashed line = ε² 0.14"
  )
)
```

```{r plot-top25-h-pvalue}
plot_top_features_h_pvalue(
  time_feature_stats = time_feature_stats,
  n_features = 25,
  file = file.path(OUT$stats, "top25_h_pvalue.png"),
  title = "Top 25 features — H statistic and p-value",
  subtitle = "Ranked by ε²"
)
```

# Selected heatmaps

```{r heatmap-top10-plus-extras}
feats_top10 <- select_top_plus_extras(time_feature_stats, n = 10)
plot_image_heatmap(
  scaled_mat = scale_features(feature_matrix(image_means, feats_top10)),
  row_ann = row_annotation,
  ann_colors = ann_colors,
  title = "Top 10 features + skeleton + GFAP intensity",
  file = file.path(OUT$heatmaps, "top10_plus_extras.png"),
  fontsize_row = 10,
  fontsize_col = 10
)
```

```{r heatmap-cosine-dedup-core}
feats_dedup_core <- dedup_features_by_cosine(
  ranked_stats = core_time_stats,
  feature_mat = mean_matrix,
  candidate_features = core_time_stats$Feature[seq_len(min(20, nrow(core_time_stats)))],
  cosine_threshold = 0.85,
  force_include = SKELETON_INTENSITY_FEATURES
)
plot_image_heatmap(
  scaled_mat = scale_features(feature_matrix(image_means, feats_dedup_core)),
  row_ann = row_annotation,
  ann_colors = ann_colors,
  title = "Cosine-deduped core features (|r| < 0.85) + skeleton + intensity",
  file = file.path(OUT$heatmaps, "cosine_dedup_core.png"),
  fontsize_row = 10,
  fontsize_col = 10
)
```

```{r heatmap-skeleton-intensity}
plot_image_heatmap(
  scaled_mat = scale_features(feature_matrix(image_means, SKELETON_INTENSITY_FEATURES)),
  row_ann = row_annotation,
  ann_colors = ann_colors,
  title = "Skeleton + GFAP integrated intensity",
  file = file.path(OUT$heatmaps, "skeleton_intensity.png"),
  fontsize_row = 10,
  fontsize_col = 10
)
```

```{r heatmap-final}
feats_final <- c(
  "AreaShape_MeanRadius",
  "AreaShape_FormFactor",
  "ObjectSkeleton_TotalObjectSkeletonLength_MorphologicalSkeleton",
  "ObjectSkeleton_NumberNonTrunkBranches_MorphologicalSkeleton",
  "Intensity_IntegratedIntensity_GFAP_Images"
)
mat_final <- log_transform_intensity_columns(feature_matrix(image_means, feats_final))
plot_image_heatmap(
  scaled_mat = scale_features(mat_final),
  row_ann = row_annotation,
  ann_colors = ann_colors,
  title = "Final panel (MeanRadius, FormFactor, skeleton, log10 GFAP intensity)",
  file = file.path(OUT$heatmaps, "final.png"),
  fontsize_row = 10,
  fontsize_col = 10
)
```

# Boxplots

Union of top-10-plus-extras and cosine-deduped core features, ordered by ε².

```{r boxplots-final}
all_stats <- bind_rows(time_feature_stats, core_time_stats) %>%
  distinct(Feature, .keep_all = TRUE)

boxplot_feats <- all_stats %>%
  filter(Feature %in% c(feats_top10, feats_dedup_core)) %>%
  arrange(desc(epsilon_squared)) %>%
  pull(Feature)

plot_time_boxplots(
  image_means = image_means,
  top_features = boxplot_feats[1:8],
  time_colors = TIME_COLORS,
  title = "Selected features 1–8",
  subtitle = "Each dot = one image; ordered by Kruskal–Wallis ε²",
  file = file.path(OUT$boxplots, "boxplots_1_8.png"),
  width = 12,
  height = 10
)

plot_time_boxplots(
  image_means = image_means,
  top_features = boxplot_feats[9:16],
  time_colors = TIME_COLORS,
  title = "Selected features 9–16",
  subtitle = "Each dot = one image; ordered by Kruskal–Wallis ε²",
  file = file.path(OUT$boxplots, "boxplots_9_16.png"),
  width = 12,
  height = 10
)

cat("All-times plots saved to:", OUTPUT_ROOT, "\n")
```
